---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#MORE ON PCA!!! PRINCIPAL COMPONENT ANALYSIS!!! LOOK AT SESSION 5 AND NOTES FOR 10-25-19


```{r}
#fna.data <- "data/WisconsinCancer.csv"
#wisc.df <- read.csv("WisconsinCancer.csv")
#head(wisc.df)
### dang this is BIG
```

### OUR DATA IS CANCER RELATED!!! IT'S RELEVANT TO YOUR LIFE! 

### DANG THIS DATA IS A LOT!!! WHAT DO I DO!!! how do I examine data from so many patients with so many results!!! 

Here we examine r nrow(wisc_df) patient samples #this gives us the number of patients when opening the data somehow

```{r}
#table(wisc.df$diagnosis) #giving us the types of diagnosis 
```
in this data-set we have r x["M"] cancer and r["B"] non-cancer patients
##
in this data-set we have rx["_mean"]` columns 


### we dont want ID or diagnosis or thelast column, X. so we are only going to use columns 3 to 32 3:32
so no `id` `diagnosis` or `X`
```{r}
#wisc.data <- as.matrix(wisc.df[3:32])
#wisc.data
```


> Q, now many variables/features in the data are suffixed with _mean? like, how many columns? 
colnames

```{r}
#colnames(wisc.df)
### that's cool but i dont want to count! grep the rescue! grep(pattern,x, ...)
## what's our patter? it's _mean
#grep("_mean", colnames(wisc.df)) #ah ha! 3 through 12! that's 10 columns
#we can do better though 
#grep("_mean", colnames(wisc.df), value = TRUE)
##but we can get a raw number ;)
#length(grep("_mean", colnames(wisc.df), value = TRUE))
```
### anyways this is a crazy data set! let's to PCA to help us!
>Q does the data need to be scaled? #considering some of the numbers are decimals less than 1 and others are in the hundreds... yeah

## check the sd of each column
```{r}
#colMeans(wisc.data)
```

```{r}
#round(apply(wisc.data, 2, sd,), 3) #the 2 is the columns. 1 would mean the rows
```
### YES WE NEED TO SCALE THE DATA! bc the standard deviantions are crazy different

##SO LET'S DO OUR PCA
```{r}
#wisc.pr <- prcomp(wisc.data, scale=TRUE)
#summary(wisc.pr) #you can see that accumulation of the proportion, that you can rep 80% of the data with the first three PCs, and 90% with the first 7 PCs
```
```{r}
##makeing the PC plot! what component do we need? look at wisc.pr$ it has a few options
#we are looking to plot PC 1 vs PC 2 and color by M/B
#plot(wisc.pr$rotation)
#plot(wisc.pr$x)
### or do this!
#plot(wisc.pr$x[,1], wisc.pr$x[,2])
## or!!!
#plot(wisc.pr$x[,1:2] )
#but let's color it
#plot(wisc.pr$x[,1:2], col=wisc.df$diagnosis)
```

>Q  how many PCs are required to describe at least 70% of the data?

```{r}
#x<- summary(wisc.pr)
#x$importance[3,]>0.7
#which(x$importance[3,]>0.7)[1]
```
BACK TO H CLUSTERING I GUESS!!!

### scale the wisc.data data.scaled
```{r}
#data.scaled<- scale(wisc.data)
### calculate euclidean distance
#data.dist<- dist(data.scaled)
### then create the h cluster
#wisc.hclust<- hclust(data.dist, method="complete")
```

## plot it
```{r}
#plot(wisc.hclust)
#abline(wisc.hclust, col="red", lty=2, h=19)
```
### lets move on to cutting the tree. Use cutree() to cut the tree so that it has 4 clusters.
```{r}
#wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
#### that's weird. i think i did it wrong :()
#plot(wisc.hclust.clusters)
#table(wisc.hclust.clusters, wisc.df$diagnosis)
```

### Clutering PCA results!!! YAY!!!!

```{r}
#Using the minimum number of principal components required to describe at least 90%  of the variability (known to be PCs 1-7) in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward's criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.
#wisc.pr.hclust<- hclust(dist(wisc.pr$x[,1:7]), method= "ward.D2" )
#plot(wisc.pr.hclust)
```
```{r}
#grps<- cutree(wisc.pr.hclust, k=70)
#table(grps)
```

```{r}
#table(grps, wisc.df$diagnosis)
```


```{r}
#plot(wisc.pr$x[,1:2], col=grps)
```
```{r}
#plot(wisc.pr$x[,1:2], col=wisc.df$diagnosis)
```
Note the color swap here as the hclust cluster 1 is mostly "M" and cluster 2 is mostly "B" as we saw from the results of calling table(grps, diagnosis). To match things up we can turn our groups into a factor and reorder the levels so cluster 2 comes first

```{r}
#g <- as.factor(grps)
#levels(g)
#g <- relevel(g,2)
#levels(g)
#plot(wisc.pr$x[,1:2], col=g)
```

