---
title: "Class12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## set up BioConductor DESeq2

```{r eval=FALSE}
install.packages("BiocManager")
BiocManager::install()
BiocManager::install("DESeq2")
```


## Data for today's class saved into the project

```{r}
counts<- read.csv("airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata<- read.csv("airway_metadata.csv", stringsAsFactors = FALSE)
```
## let's have a peak
```{r}
head(counts)
```

### how many genes do we have 
```{r}
nrow(counts)
```

### how many experiments 
```{r}
ncol(counts)-1
```

### what are our controls ??? we check the metadata 
### let's make sure the metadata ID col matches the colnames of counts

```{r}
colnames(counts)[-1]
```

```{r}
metadata$id
```

```{r}
(colnames(counts)[-1])== (metadata$id)
```
### if you have a lage metadata you can condense it 
```{r}
all((colnames(counts)[-1])== (metadata$id))
```

## analysis compare the control to drug tested
## first we need ot access the columns of our countDaa that are control and find their mean

```{r}
metadata$dex
```
### but i want the specific columns to access!!!
```{r}
control_id<-metadata[ metadata$dex=="control", ]$id
control_id
```

```{r}
head(counts[ ,control_id])
```
#### ok we want the mean across the control rows!!!
```{r}
control_mean<-rowSums(counts[,control_id])/(length(control_id))
names(control_mean)<- counts$ensgene
```
## now for the treated
```{r}
treated_id<-metadata[ metadata$dex=="treated", ]$id
treated_id
```

```{r}
head(counts[ ,treated_id])
```
##now for the treated
```{r}
treated_mean<-rowSums(counts[,treated_id])/(length(treated_id))
names(treated_mean)<- counts$ensgene
```
### store them together!!
```{r}
meancounts<- data.frame(control_mean, treated_mean)
```
### plot control vs tested!!

```{r}
plot(meancounts$control_mean, meancounts$treated_mean)
```
### well that's no good!!! so let's plot it on a log scale (bc like, only a few have high expression! so it willbe easier to visualize)
```{r}
plot(meancounts$control_mean, meancounts$treated_mean, log="xy")
```

## points that are abberant (have differential expression) are noticably above or below the dark center line

### now lets get the fold change! if fold change is 0 then there was no difference
```{r}
meancounts$log2fc <- log2(meancounts[,"treated_mean"]/meancounts[,"control_mean"])
head(meancounts)
```
#### the next step is to EXCLUDE THE 0 VALUES!!! we DON'T WANT THEM!!! 

###
```{r}
x<- data.frame(happy=c(5,0,4,3), sad=c(0,5,5,0))
x==0
which(x==0, arr.ind =TRUE)
unique(which(x==0, arr.ind =TRUE)[,1])
```
## now let's apply this to our counts!
```{r}
#which(meancounts==0, arr.ind = TRUE)
to_rm<-unique(which(meancounts[,1:2]==0, arr.ind =TRUE)[,1])
zero_vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero_vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

### fold change time !!! we have set a threshold 
```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
sum(up.ind)
```
```{r}
head( mycounts[up.ind,] )
```

```{r}
#columns(org.Hs.eg.db)
#mycounts$symbol <- mapIds(org.Hs.eg.db,
                     #keys=row.names(mycounts), # Our genenames
                     #keytype="ENSEMBL",        # The format of our genenames
                     #column="SYMBOL",          # The new format we want to add
                     #multiVals="first")
#head(mycounts)
```
### we are gonna make our summary plot!!

### also hey remember that fold change is great but you need it to be significant! let's work with that

### DESeq Analysis 
```{r}
library(DESeq2)
```
```{r}
## ok time to do some DESeq
#set up our obejct for DESeq analysis 

dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```
```{r}
dds<- DESeq(dds)
res<-results(dds)
res
summary(res)
```

### cool let's plot it!!! 
#VOLCANO PLOT

```{r}
plot(res$log2FoldChange, res$padj)
```
### ok but that's no good! we want to see the small p values easily!!
#take the log! but the -log of res$padj
# in this way, the larger numbers are the smallest p values
```{r}
plot(res$log2FoldChange, -log(res$padj))
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.05), col="gray", lty=2)
```
#### let's color it 
```{r}
mycols<- rep("grey", length(res$padj))
mycols[abs(res$log2FoldChange)>2]= "blue"
mycols[ res$padj > 0.05 ]="red"
plot(res$log2FoldChange, -log(res$padj), col=mycols)

```

